Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-16 _tex/paper2.tex
Built site for gh-pages

Introduction

@@

Thurstone's theory

@@

The two critical issues in CJ literature

@@

\section{The Case V and the statistical analysis of CJ
data}\label{sec-theory-issue1}

@@

\subsection{The assumption of equal dispersions between
stimuli}\label{sec-theory-issue1a}

@@

\subsection{The assumption of zero correlation between
stimuli}\label{sec-theory-issue1b}

@@

\section{The disconnect between trait measurement and hypothesis
testing}\label{sec-theory-issue2}

@@

Updating the theoretical and statistical model

@@

\section{The theoretical model}\label{sec-theory-theoretical}

@@

\subsection{The population model}\label{sec-theory-theoretical_P}

@@

\subsection{The sample-comparison
model}\label{sec-theory-theoretical_SC}

@@

\section{From theory to statistics}\label{sec-theory-statistics}

@@

Discussion

\section{Findings}\label{sec-discuss-finding}

\section{Limitations and further
research}\label{sec-discuss-limitations}

Conclusion

@@

Declarations

@@

Appendix

This section introduces fundamental statistical and causal inference
concepts necessary for understanding the core theoretical principles
described in Section~\ref{sec-theory}. It does not, however, offer a
comprehensive overview of causal inference methods. Readers seeking more
in-depth understanding may wish to explore introductory papers such as
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find it helpful to consult
introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
\citet{McElreath_2020}. For more advanced study, readers may refer to
seminal intermediate papers such as \citet{Neyman_et_al_1923},
\citet{Rubin_1974}, \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009},
as well as books such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
and \citet{Hernan_et_al_2020}.

\section{Empirical research and randomized
experiments}\label{sec-appendix-A}

Empirical research uses evidence from observation and experimentation to
address real-world challenges. In this context, researchers typically
formulate their research questions as \emph{estimands} or \emph{targets
of inference}, i.e., the specific quantities they seek to determine
\citep{Everitt_et_al_2010}. For instance, researchers might be
interested in answering the following question: ``To what extent do
different teaching methods \((T)\) influence students' ability to
produce high-quality written texts \((Y)\)?'' To investigate this,
researchers could randomly assign students to two groups, each exposed
to a different teaching method \((T_{i} = \{1,2\})\). Then, they would
perform pairwise comparisons, generating a dichotomous outcome
\((Y_{i} = \{0,1\})\) showing which student exhibits more of the
ability. In this scenario, the research question can be rephrased as the
estimand, ``\emph{On average}, is there a difference in the ability to
produce high-quality written texts between the two groups of students?''
and this estimand can be mathematically represented by the random
quantity \(E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), where \(E[\cdot]\)
denotes the expected value.

Researchers would then proceed to identify the estimands.
\emph{Identification} is concerned with determining whether an estimator
can accurately compute an estimand based on its assumptions, regardless
of random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
\emph{estimator} is a method or function that transforms data into an
estimate \citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand and are derived through \emph{estimation},
which refers to the process of integrating data with an estimator
\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow} provides a
visual representation of the process of transitioning from estimands to
estimates.

@@

While numerous methods can approximate an estimand, researchers
prioritize estimators with desirable properties that ensure the accuracy
of estimates. For instance, the Z-test is an estimator known for its
effectiveness in comparing groups' proportions, yielding accurate
estimates when the underlying assumptions of the statistic are met
\citep{Kanji_2006}. If this is the case, the Z-test is expressed as a
signal-to-noise statistic
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
as the difference between the groups' sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y_{i}| T_{i}=1]\) and \(E[Y_{i}| T_{i}=2]\), respectively. The
noise, represented by \(\hat{s}_{p}\), is defined as the unpooled sample
variability observed between the two groups.

@@

\section{Identification under causal inference}\label{sec-appendix-B}

@@

\emph{Counterfactuals} are hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The structural approach to
causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
framework for defining counterfactuals. For instance, in the scenario
described in Section~\ref{sec-appendix-A}, the approach begins by
defining the \emph{individual causal effect} (ICE) as the difference
between each student's potential outcomes:
\(\tau_{i} = Y_{i}|do(T_{i}=1) - Y_{i}|do(T_{i}=2)\). Here,
\(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i}|do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i}|do(T_{i}=1)\) represents the
potential outcome under intervention \(T_{i}=2\). Note that if a student
is assigned to intervention \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) becomes a counterfactual, as it is no longer observed nor
observable. To address the challenge of unobserved counterfactuals, the
approach extends the ICE to the \emph{average causal effect} (ACE):
\(\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]\),
representing the average difference between observed potential outcomes
and their counterfactual counterparts.

@@

For instance, if researchers are unable to conduct the randomized
experiments described in Section~\ref{sec-appendix-A} and instead rely
on observational data, they can still estimate the ACE provided a
variable \(X\), such as the socio-economic status of the school, blocks
all non-causal paths between the teaching method \(T\) to the outcome
\(Y\) without opening any new ones. Under these conditions, the ACE can
be estimated from associational quantities as \(\tau =\)
\(E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)] =\)
\(E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]\),
where \(E_{X}[\cdot]\) represents the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}. Notably, the approach extends the ACE for a
continuous variable \(T\) as \(\tau =\) \(E[Y_{i}|do(T_{i}=t)] =\)
\(d E_{X}\left[ E[Y_{i}| T_{i}=t, X]\right]/ dt\), ensuring broad
applicability across different causal scenarios
\citep[pp.~45]{Neal_2020}.

\section{SCMs and DAGs}\label{sec-appendix-C}

\newcommand{\dsep}{\perp\!\!\!\perp}
\newcommand{\ndsep}{\not\!\perp\!\!\!\perp}

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Assuming the depicted causal structure is correct, SCMs and DAGs serve
as \emph{conceptual (theoretical) models} that help researchers
determine which statistical models can identify an estimand, thus
enabling valid causal inference \citep{McElreath_2020}. They achieve
this by representing various causal structures using only five
fundamental building blocks, regardless of complexity
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components
\citep{McElreath_2020}. Furthermore, SCMs and DAGs can depict causal
relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
requiring detailed specifications of the functional relating the
variables, their parameters, or their data types
\citep{Pearl_et_al_2016}. Figure~\ref{fig-IEflow} shows how theoretical
models guide the inference process.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display these five fundamental building blocks. The left panels of the
figures show the formal mathematical models, represented by the SCMs,
defined in terms of a set of \emph{endogenous} variables
\(V=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
figures display these DAGs. A DAG is a graph consisting of nodes
connected by edges, where the nodes represent random variables. The term
\emph{directed} means that the edges extend from one node to another,
with arrows indicating the direction of causal influence. The term
\emph{acyclic} implies that the causal influences do not form loops,
ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Finally, the arrows in the graphs indicate
the expected direction of causal influences between these variables.
While DAGs often omit exogenous variables for simplicity (the
\emph{standard} representation), including these variables is
advantageous (the \emph{magnified} representation shown in the figures),
as their inclusion helps to identify potential issues related to
conditioning and confounding \citep{Cinelli_et_al_2020}.

@@

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb1.png}

@@

\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb2.png}

@@

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb3.png}

@@

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb4.png}

@@

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb5.png}

@@

A careful examination of the figures highlights the theoretical
assumptions underlying these building blocks. Figures \ref{fig-scm_bb1}
and \ref{fig-mdag_bb1} depict two unconnected nodes, representing a
scenario where variables \(X_{1}\) and \(X_{3}\) are not causally
related. Figures \ref{fig-scm_bb2} and \ref{fig-mdag_bb2} illustrate two
connected nodes, representing a scenario where a \emph{parent} node
\(X_{1}\) exerts a causal influence on a \emph{child} node \(X_{3}\). In
this setup, \(X_{3}\) is considered a \emph{descendant} of \(X_{1}\).
Additionally, \(X_{1}\) and \(X_{3}\) are described as \emph{adjacent}
because an edge directly connects them. Figures \ref{fig-scm_bb3} and
\ref{fig-mdag_bb3} depict a \emph{chain}, where \(X_{1}\) influences
\(X_{2}\), and \(X_{2}\) influences \(X_{3}\). In this configuration,
\(X_{1}\) is a parent node of \(X_{2}\), which is a parent node of
\(X_{3}\). This structure creates a \emph{directed path} between
\(X_{1}\) and \(X_{3}\). Consequently, \(X_{1}\) is an \emph{ancestor}
of \(X_{3}\), and \(X_{2}\) fully \emph{mediates} the relationship
between the two. Figures \ref{fig-scm_bb4} and \ref{fig-mdag_bb4}
illustrate a \emph{fork}, where variables \(X_{1}\) and \(X_{3}\) are
both influenced by \(X_{2}\). Here, \(X_{2}\) is a parent node that
\emph{confounds} the relationship between \(X_{1}\) and \(X_{3}\).
Finally, Figures \ref{fig-scm_bb5} and \ref{fig-mdag_bb5} show a
\emph{collider}, where variables \(X_{1}\) and \(X_{3}\) are concurrent
causes of \(X_{2}\). In this configuration, \(X_{1}\) and \(X_{3}\) are
not causally related to each other but both influence \(X_{2}\) (an
\emph{inmorality}). Notably, in all SCMs, the errors are assumed to be
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X1} \perp\!\!\!\perp e_{X2}\),
\(e_{X1} \perp\!\!\!\perp e_{X3}\), and
\(e_{X2} \perp\!\!\!\perp e_{X3}\).

Researchers can use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. Figures \ref{fig-scm_example1}
and \ref{fig-mdag_example1} depict the plausible causal structure for
this example, assuming that the variable \(X\) (socio-economic status of
the school) acts as a confounder in the relationship between the
teaching method \(T\) and the outcome \(Y\). The figures show several
descendant, such as \(X \rightarrow T\), \(X \rightarrow Y\), and
\(T \rightarrow Y\), and also highlight multiple pairs of unconnected
nodes, evident from the relationships \(e_{T} \perp\!\!\!\perp e_{X}\),
\(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
Additional, the figures depict one fork, \(X \rightarrow \{T, Y\}\), and
two colliders: \(\{X, e_{T}\} \rightarrow T\) and
\(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1.png}

@@

\section{The probabilistic implications of SCMs and
DAGs}\label{sec-appendix-D}

In addition to their graphical capabilities, SCMs and DAGs can encode
probabilistic information about causal structures. They achieve this
encoding by relying on three fundamental assumptions: the local Markov,
the minimality, and the causal edges assumption. The \emph{local Markov
assumption} encodes probabilistic independencies between variables by
declaring that nodes in a graph are independent of all its
non-descendants, given its parents \citep{Neal_2020}. Meanwhile, the
\emph{minimality assumption} encodes probabilistic dependencies among
variables by stating that every pair of adjacent nodes exhibits a
dependency \citep{Neal_2020}. Finally, the \emph{causal edges
assumption} encodes causal relationships between variables by declaring
that each parent node acts as a direct cause of its children
\citep{Neal_2020}. Figure~\ref{fig-ACflow} illustrates how these
assumptions influence the statistical and causal interpretations of
graphs.

@@

The encoding of probabilistic information in SCMs and DAGs facilitates
the efficient representation of the joint distribution of variables
\citep[pp.~29]{Pearl_et_al_2016}. A notable implication of the
assumptions underlying this encoding is that any conceptual model
described by an SCM and DAG can express the joint distribution of
variables as a product of conditional probability distributions (CPDs)
in the form \(P(child \: | \: parents)\). This property is formally
known as the \emph{Bayesian Network factorization}
(Equation~\ref{eq-net_factor})
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this factorization, \(pa(X_{i})\) denotes the set of variables that
are the parents of \(X_{i}\).

@@

Ultimately, this encoding enables researchers with conceptual
(theoretical) knowledge in the form of an SCM and DAG to predict
patterns of (in)dependencies in the data. As highlighted by
\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
structure of these conceptual models without requiring the quantitative
details of the equations or the distributions of the errors. Moreover,
once researchers observe empirical data, the patterns of
(in)dependencies in this data can provide significant insights into the
validity of the proposed conceptual model.

The concept of predicting patterns of (in)dependencies in the data is
more clearly illustrated through the five fundamental building blocks
outlined in Section~\ref{sec-appendix-C}.

>>>>

\section{Introduction}\label{sec-introduction}

@@

\section{Thurstone's theory}\label{sec-thurstone_theory}

@@

\section{The two critical issues in CJ
literature}\label{sec-theory-issues}

@@

\subsection{The Case V and the statistical analysis of CJ
data}\label{sec-theory-issue1}

@@

\subsubsection{The assumption of equal dispersions between
stimuli}\label{sec-theory-issue1a}

@@

\subsubsection{The assumption of zero correlation between
stimuli}\label{sec-theory-issue1b}

@@

\subsection{The disconnect between trait measurement and hypothesis
testing}\label{sec-theory-issue2}

@@

\section{Updating the theoretical and statistical
model}\label{sec-theory}

@@

\subsection{The theoretical model}\label{sec-theory-theoretical}

@@

\subsubsection{The population model}\label{sec-theory-theoretical_P}

@@

\subsubsection{The sample-comparison
model}\label{sec-theory-theoretical_SC}

@@

\subsection{From theory to statistics}\label{sec-theory-statistics}

@@

\section{Discussion}\label{sec-discuss}

\subsection{Findings}\label{sec-discuss-finding}

\subsection{Limitations and further
research}\label{sec-discuss-limitations}

\section{Conclusion}\label{sec-conclusion}

@@

\section*{Declarations}\label{declarations}
\addcontentsline{toc}{section}{Declarations}

@@

\section{Appendix}\label{sec-appendix}

This section introduces fundamental statistical and causal inference
concepts necessary for understanding the core theoretical principles
described in this document. It does not, however, offer a comprehensive
overview of causal inference methods. Readers seeking more in-depth
understanding may wish to explore introductory papers such as
\citet{Pearl_2010}, \citet{Rohrer_2018}, \citet{Pearl_2019}, and
\citet{Cinelli_et_al_2020}. They may also find it helpful to consult
introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}, and
\citet{McElreath_2020}. For more advanced study, readers may refer to
seminal intermediate papers such as \citet{Neyman_et_al_1923},
\citet{Rubin_1974}, \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009},
as well as books such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014},
and \citet{Hernan_et_al_2020}.

\subsection{Empirical research and randomized
experiments}\label{sec-appendix-A}

Empirical research uses evidence from observation and experimentation to
address real-world challenges. In this context, researchers typically
formulate their research questions as \emph{estimands} or \emph{targets
of inference}, i.e., the specific quantities they seek to determine
\citep{Everitt_et_al_2010}. For instance, researchers might be
interested in answering the following question: ``To what extent do
different teaching methods \((T)\) influence students' ability to
produce high-quality written texts \((Y)\)?'' To investigate this,
researchers could randomly assign students to two groups, each exposed
to a different teaching method \((T_{i} = \{1,2\})\). Then, they would
perform pairwise comparisons, generating a dichotomous outcome
\((Y_{i} = \{0,1\})\) showing which student exhibits more of the
ability. In this scenario, the research question can be rephrased as the
estimand, ``\emph{On average}, is there a difference in the ability to
produce high-quality written texts between the two groups of students?''
and this estimand can be mathematically represented by the random
quantity \(E[Y_{i} \: | \: T_{i}=1] - E[Y_{i} \: | \: T_{i}=2]\), where
\(E[\cdot]\) denotes the expected value.

Researchers would then proceed to identify the estimands.
\emph{Identification} is concerned with determining whether an estimator
can accurately compute an estimand based on its assumptions, regardless
of random variability \citep[pp.~4]{Schuessler_et_al_2023}. An
\emph{estimator} is a method or function that transforms data into an
estimate \citep{Neal_2020}. \emph{Estimates} are numerical values that
approximate the estimand. They are derived through \emph{estimation},
which refers to the process of integrating data with an estimator
\citep{Everitt_et_al_2010}. The Identification-Estimation flowchart
\citep{McElreath_2020, Neal_2020} in Figure~\ref{fig-IEflow} provides a
visual representation of the process of transitioning from estimands to
estimates.

@@

While numerous methods can approximate an estimand, researchers
prioritize estimators with desirable identification properties that
ensure the accuracy of estimates. For instance, the Z-test is an
estimator known for its effectiveness when comparing groups'
proportions, yielding accurate estimates when the underlying assumptions
of the statistic are met \citep{Kanji_2006}. If this is the case, the
Z-test is expressed as a signal-to-noise statistic
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is defined
as the difference between the groups' sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y_{i} \: | \: T_{i}=1]\) and \(E[Y_{i} \: | \: T_{i}=2]\),
respectively. The noise, represented by \(\hat{s}_{p}\), is defined as
the unpooled sample variability observed between the two groups.

@@

\subsection{Identification under causal inference}\label{sec-appendix-B}

@@

\emph{Counterfactuals} are hypothetical scenarios that are
\emph{contrary to fact}, where alternative outcomes resulting from a
given cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}. The structural approach to
causal inference \citep{Pearl_2009, Pearl_et_al_2016} provides a formal
framework for defining counterfactuals. For instance, in the scenario
described in Section~\ref{sec-appendix-A}, the approach begins by
defining the \emph{individual causal effect} (ICE) as the difference
between each student's potential outcomes:
\(\tau_{i} = Y_{i} \: | \: do(T_{i}=1) - Y_{i} \: | \: do(T_{i}=2)\).
Here, \(do(T_{i}=t)\) represents the intervention operator,
\(Y_{i} \: | \: do(T_{i}=1)\) represents the potential outcome under
intervention \(T_{i}=1\), and \(Y_{i} \: | \: do(T_{i}=1)\) represents
the potential outcome under intervention \(T_{i}=2\). Note that if a
student is assigned to intervention \(T_{i}=1\), the potential outcome
under \(T_{i}=2\) becomes a counterfactual, as it is no longer observed
nor observable. To address the challenge of unobserved counterfactuals,
the approach extends the ICE to the \emph{average causal effect} (ACE):
\(\tau = E[\tau_{i}] = E[Y_{i} \: | \: do(T_{i}=1)]- E[Y_{i} \: | \: do(T_{i}=2)]\),
representing the average difference between observed potential outcomes
and their counterfactual counterparts.

@@

For instance, if researchers are unable to conduct the randomized
experiments described in Section~\ref{sec-appendix-A} and must rely on
observational data, they can still estimate the ACE, provided that a
variable \(X\), such as the socio-economic status of the school, blocks
all non-causal paths between the teaching method \(T\) and the outcome
\(Y\) without opening any new ones. Under these conditions, the ACE can
be estimated from the \emph{conditional average causal effect} (CACE),
defined by \(E[Y_{i} \: | \: T_{i}=t, X]\). Using the CACE, the ACE can
be expressed as \(\tau =\)
\(E[Y_{i} \: | \: do(T_{i}=1)] - E[Y_{i} \: | \: do(T_{i}=2)] =\)
\(E_{X}\left[ E[Y_{i} \: | \: T_{i}=1, X] - E[Y_{i} \: | \: T_{i}=2, X] \right]\),
where \(E_{X}[\cdot]\) represents the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}. Notably, the approach extends the ACE for a
continuous variable \(T\) as \(\tau =\)
\(E[Y_{i} \: | \: do(T_{i}=t)] =\)
\(d E_{X}\left[ E[Y_{i} \: | \: T_{i}=t, X]\right]/ dt\), ensuring broad
applicability across different causal scenarios
\citep[pp.~45]{Neal_2020}.

\subsection{SCMs and DAGs}\label{sec-appendix-C}

\newcommand{\dsep}{\perp\!\!\!\perp}

The structural approach to causal inference uses SCMs and DAGs to
formally and graphically represent the presumed causal structure
underlying the ACE
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
Essentially, SCMs and DAGs serve as \emph{conceptual (theoretical)
models} that help researchers determine which statistical models can
identify an estimand, assuming the depicted causal structure is correct,
thus enabling valid causal inference \citep{McElreath_2020}. They
achieve this by representing various causal structures, regardless of
complexity, using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature allows researchers to
decompose complex structures into manageable components, facilitating
their analysis \citep{McElreath_2020}. Furthermore, SCMs and DAGs depict
causal relationships in a non-parametric and fully interactive way. This
flexibility enables feasible ACE identification strategies without
requiring detailed specifications of functional relationships relating
the variables, their parameters, or their data types
\citep{Pearl_et_al_2016}. Figure~\ref{fig-IEflow} shows the role of
theoretical models in the inference process.

Figures \ref{fig-dags_scms1}, \ref{fig-dags_scms2},
\ref{fig-dags_scms3}, \ref{fig-dags_scms4}, and \ref{fig-dags_scms5}
display these five fundamental building blocks. The left panels of the
figures show the formal mathematical models, represented by the SCMs,
defined in terms of a set of \emph{endogenous} variables
\(X=\{X_{1},X_{2},X_{3}\}\), a set of \emph{exogenous} variables
\(E=\{e_{X1},e_{X2},e_{X3}\}\), and a set of functions
\(F=\{f_{X1},f_{X2},f_{X3}\}\) \citep{Pearl_2009, Cinelli_et_al_2020}.
Endogenous variables are those whose causal mechanisms a researcher
chooses to model \citep{Neal_2020}. In contrast, exogenous variables
represent \emph{errors} or \emph{disturbances} arising from omitted
factors that the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence of the
variables and the symbol `\(\perp\!\!\!\perp\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. The right panels of the
figures display these DAGs. A DAG is a graph consisting of nodes
connected by edges, where the nodes represent random variables. The term
\emph{directed} means that the edges extend from one node to another,
with arrows indicating the direction of causal influence. The term
\emph{acyclic} implies that the causal influences do not form loops,
ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. Although DAGs typically omit exogenous
variables for simplicity (the \emph{standard representation}), including
these variables, as demonstrated in the \emph{magnified representation}
in the figures, can be advantageous. Their inclusion helps to highlight
potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

@@

\includegraphics[width=0.54\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb1.png}

@@

\includegraphics[width=0.54\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb2.png}

@@

\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb3.png}

@@

\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb4.png}

@@

\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_bb5.png}

@@

A careful examination of the figures highlights the theoretical
assumptions underlying these building blocks. SCM \ref{fig-scm_bb1} and
DAG \ref{fig-mdag_bb1} depict two unconnected nodes, representing a
scenario where variables \(X_{1}\) and \(X_{3}\) are independent or not
causally related. SCM \ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}
illustrate two connected nodes, representing a scenario where a
\emph{parent} node \(X_{1}\) exerts a causal influence on a \emph{child}
node \(X_{3}\). In this setup, \(X_{3}\) is considered a
\emph{descendant} of \(X_{1}\). Additionally, \(X_{1}\) and \(X_{3}\)
are described as \emph{adjacent} because an edge directly connects them.
SCM \ref{fig-scm_bb3} and DAG \ref{fig-mdag_bb3} depict a \emph{chain},
where \(X_{1}\) influences \(X_{2}\), and \(X_{2}\) influences
\(X_{3}\). In this configuration, \(X_{1}\) is a parent node of
\(X_{2}\), which is a parent node of \(X_{3}\). This structure creates a
\emph{directed path} between \(X_{1}\) and \(X_{3}\). Consequently,
\(X_{1}\) is an \emph{ancestor} of \(X_{3}\), and \(X_{2}\) fully
\emph{mediates} the relationship between the two. SCM \ref{fig-scm_bb4}
and DAG \ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables
\(X_{1}\) and \(X_{3}\) are both influenced by \(X_{2}\). Here,
\(X_{2}\) is a parent node that \emph{confounds} the relationship
between \(X_{1}\) and \(X_{3}\). Finally, SCM \ref{fig-scm_bb5} and DAG
\ref{fig-mdag_bb5} show a \emph{collider}, where variables \(X_{1}\) and
\(X_{3}\) are concurrent causes of \(X_{2}\). In this configuration,
\(X_{1}\) and \(X_{3}\) are not causally related to each other but both
influence \(X_{2}\) (an \emph{inmorality}). Notably, all building blocks
assume the errors are independent of each other and from all other
variables in the graph, as evidenced by the pairwise relations
\(e_{X1} \perp\!\!\!\perp e_{X2}\), \(e_{X1} \perp\!\!\!\perp e_{X3}\),
and \(e_{X2} \perp\!\!\!\perp e_{X3}\).

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) acts as a confounder in the relationship between the
teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

@@

\includegraphics[width=0.65\linewidth,height=\textheight,keepaspectratio]{images/png/mdag_example1.png}

@@

\subsection{The probabilistic and causal implications of SCMs and
DAGs}\label{sec-appendix-D}

Beyond their graphical capabilities, SCMs and DAGs can encode the
probabilistic information embedded within a causal structure. They
achieve this encoding by relying on three fundamental assumptions: the
local Markov, the minimality, the causal edges and the modularity
assumptions. The \emph{local Markov assumption} encodes probabilistic
independencies between variables by declaring that nodes in a graph are
independent of all its non-descendants, given its parents
\citep[pp.~20]{Neal_2020}. Meanwhile, the \emph{minimality assumption}
encodes probabilistic dependencies among variables by stating that every
pair of adjacent nodes exhibits a dependency \citep[pp.~21]{Neal_2020}.
Finally, the \emph{causal edges assumption} encodes causal relationships
between variables by declaring that each parent node acts as a direct
cause of its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
illustrates how these assumptions influence the statistical and causal
interpretations of graphs.

@@

The encoding of probabilistic information in SCMs and DAGs facilitates
the efficient representation of the joint distribution of variables
\citep[pp.~29]{Pearl_et_al_2016}. A notable implication of the
assumptions underlying the probabilistic encoding is that any conceptual
model described by an SCM and DAG can express the joint distribution of
variables as a product of conditional probability distributions (CPDs)
in the form \(P(child \: | \: parents)\). This property is formally
known as the \emph{Bayesian Network factorization} (BNF)
\citetext{\citealp[pp.~29]{Pearl_et_al_2016}; \citealp[pp.~21]{Neal_2020}}.
In this factorization, depicted in Equation~\ref{eq-net_factor},
\(pa(X_{i})\) denotes the set of variables that are the parents of
\(X_{i}\).

@@

This encoding enables researchers with conceptual (theoretical)
knowledge in the form of an SCM and DAG to predict patterns of
(in)dependencies in the data. As highlighted by
\citet[pp.~35]{Pearl_et_al_2016}, these predictions depend solely on the
structure of these conceptual models without requiring the quantitative
details of the equations or the distributions of the errors. Moreover,
once researchers observe empirical data, the patterns of
(in)dependencies in this data can provide significant insights into the
validity of the proposed conceptual model. Moreover, it allows to
express

The five fundamental building blocks described in
Section~\ref{sec-appendix-C} illustrate how to predict (in)dependencies
in data. For instance, applying the BNF to the causal structure shown in
the SCM \ref{fig-scm_bb1} and DAG \ref{fig-mdag_bb1} supports the
theoretical assumption that the observed variables \(X_{1}\) and
\(X_{3}\) are unconditionally independent
(\(X_{1} \perp\!\!\!\perp X_{3}\)). This conclusion follows from
expressing their joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})\), which defines probabilistic
independence \citep[pp.~24]{Neal_2020}. Conversely, when \(X_{3}\) is
unconditionally dependent on \(X_{1}\), as depicted in the SCM
\ref{fig-scm_bb2} and DAG \ref{fig-mdag_bb2}, the BNF express their
joint probability distribution as
\(P(X_{1}, X_{3}) = P(X_{3} \: | \: X_{1}) P(X_{1})\). Notably, these
descriptions demonstrate the clear correspondence between the structural
equations illustrated in the figures of Section~\ref{sec-appendix-C} and
the CPDs.
